Engineering infinite-range SU($n$) interactions with spin-orbit-coupled fermions in an optical lattice

### 2021-09-30 sun_phases/spin_model_paper/main.tex
updating submitted versions of spin model and tomography papers

\date{\today}

\begin{abstract}
We study multilevel fermions in an optical lattice described by the Hubbard model with on site SU($n$)-symmetric interactions.
We show that in an appropriate parameter regime this system can be mapped onto a spin model with all-to-all SU($n$)-symmetric couplings.
Raman pulses that address internal spin states modify the atomic dispersion relation and induce spin-orbit coupling, which can act as a synthetic inhomogeneous magnetic field that competes with the SU($n$) exchange interactions.
We investigate the mean-field dynamical phase diagram of the resulting model as a function of $n$ and different initial configurations that are accessible with Raman pulses.
Consistent with previous studies for $n=2$, we find that for some initial states the spin model exhibits two distinct dynamical phases that obey simple scaling relations with $n$.
Moreover, for $n>2$ we find that dynamical behavior can be highly sensitive to initial intra-spin coherences.
Our predictions are readily testable in current experiments with ultracold alkalike-earth(-like) atoms.
\end{abstract}

@@

SU($n$) symmetries play an important role in physics.
Underpinning much of high energy physics, the SU($n$) gauge theory known as Yang-Mills theory is central to our understanding of the electroweak and strong forces.
Extensions of Yang-Mills and SU($n$) symmetry feature in the most well-studied examples of holographic duality \cite{maldacena1999largen} and the connection between entanglement and gravity \cite{ryu2006holographic} through the anti-de Sitter/conformal field theory (AdS/CFT) correspondence.
In a condensed matter setting, SU($2$) appears ubiquitously as a symmetry of the Hubbard model, with important consequences for the study of quantum magnetism and high temperature superconductivity \cite{lee2006doping}.
The extension of SU($2$) Hubbard and spin models to SU($n$) has led to predictions of exotic phases of matter such as valence bond solids \cite{read1989valencebond, rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and chiral spin liquids \cite{hermele2009mott, hermele2011topological, chen2016syntheticgaugefield, nataf2016chiral}, as well as the potential to perform universal topological quantum computation \cite{freedman2004class, nayak2008nonabelian} and other phenomena \cite{nataf2014exact, nataf2016exact}.
Furthermore, disordered SU($n$) spin models have opened analytically tractable avenues for studying quantum chaos and information scrambling \cite{sachdev1993gapless}.

@@

In the spirit of quantum simulation, further investigations in controlled settings will play an important role in understanding the consequence of SU($n$) symmetries for fundamental questions in physics, as well as their practical use in technological applications.
For example, SU($2$)-symmetric spin interactions can be harnessed to develop quantum sensors that surpass classical limits on measurement precision \cite{he2019engineering, perlin2020spin}.
The prospect of similarly exploiting more general SU($n$) symmetries to achieve a technological advantage is still an unexplored avenue of research with untapped potential.

In this work, we consider an experimentally relevant and theoretically tractable regime of the SU($n$) Hubbard model, highlighting differences and similarities with the more familiar case of SU($2$).
Working at ultracold temperatures and unit spatial filling (one atom per lattice site), we begin by mapping the SU($n$) Hubbard model onto a multilevel spin model with all-to-all SU($n$)-symmetric interactions in Section \ref{sec:spin_model}.
In Section \ref{sec:controls} we consider the use of control fields to address nuclear spins, finding a simple three-laser driving scheme that allows for the preparation of interesting states with nontrivial intra-spin correlations when $n>2$.
We consider the effect of spin-orbit coupling (SOC) induced by control fields in Section \ref{sec:SOC}, finding in particular that the weak-SOC limit generally gives rise to a (synthetic) inhomogeneous magnetic field, extending previously known results to $n>2$ \cite{mancini2015observation, wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
Finally, we combine these ingredients to examine mean-field dynamical behaviors of the SU($n$) spin model in Section \ref{sec:mean_field}, finding that:
\begin{enumerate*}
\item long-time-averaged observables obey simple scaling relations with the spin dimension $n$, exhibiting (for spin-polarized initial states) dynamical ferromagnetic and dynamical paramegnetic phases, as previously seen for the case of $n=2$ \cite{smale2019observation, lewis-swan2021cavityqed}, and
\item for $n>2$ the long-time dynamics can be highly sensitive to the intra-spin coherences of the initial state.
\end{enumerate*}
We conclude and discuss future directions in Section \ref{sec:conclusions}.

@@

Here we derive a collective SU($n$) spin model for a system of ultracold alkaline-earth(-like) atoms trapped in an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by the single-body kinetic and two-body interaction Hamiltonians
\begin{align}
  \H_{\t{kin}}
  &= -J \sum_{\bk{j,j'},\mu} \c_{j\mu}^\dag \c_{j'\mu} + \t{h.c.},
  \label{eq:H_kin_FH} \\
  \H_{\t{int}}
  &= \f{U}{2} \sum_{j,\mu,\nu}
  \c_{j\mu}^\dag \c_{j\mu} \c_{j\nu}^\dag \c_{j\nu},
  \label{eq:H_int_FH}
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $\c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude (for simplicity assumed to be the same in all directions); and $U$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which is a good approximation for a sufficiently deep lattice, namely when $J\lesssim E_{\t{R}}$, where $E_{\t{R}}$ is the atom recoil energy.
For simplicity, we now assume a one-dimensional periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$ (in units with lattice spacing $a=1$), $\c_{j\mu} = \frac1{\sqrt{L}} \sum_k e^{-\i q\cdot j} \c_{q\mu}$, finding that
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \c_{q\mu}^\dag \c_{q\mu}, \\
  \H_{\t{int}}
  &= \f{u}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  \c_{k\mu}^\dag \c_{\ell\mu} \c_{p\nu}^\dag \c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $u\equiv U\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum).

If the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, then the mode-changing collisions in $\H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (which we assume for the remainder of this work)\footnote{Note that the frozen-mode approximation neglects correlated momentum-hopping terms of the form $\c_{\pi-p,\mu}^\dag \c_{\pi-q,\mu} \c_{p\nu}^\dag \c_{q\nu}$, which conserve both momentum and energy.
We defer a careful treatment of these terms to future work, noting only that they vanish on the manifold of permutationally symmetric spin states with one atom per lattice site, and that the frozen-mode approximation is benchmarked in Refs.~\cite{he2019engineering, smale2019observation} and Appendix \ref{sec:benchmarking}.}.
The terms with $k=\ell$ and $p=q$ are $\frac{u}{2N} \sum \c_{\ell\mu}^\dag \c_{\ell\mu} \c_{q\nu}^\dag \c_{q\nu} = \frac12 N u$, which is a constant  energy shift that we can freely neglect.
Defining the spin operators $\s_{\mu\nu q}\equiv \c_{q\mu}^\dag \c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians are
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \s_{\mu\mu q},
  \label{H_kin_start} \\
  \H_{\t{int}}
  &= -\f{u}{2N} \sum_{p,q,\mu,\nu} \s_{\mu\nu p} \s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case we can simply treat our system as $N$ distinguishable $n$-level quantum spins at ``lattice sites'' $p,q$.
Note that the ``kinetic'' terms of this spin model ($\H_{\t{kin}}$) are proportional to the identity operator, contributing an overall shift in energy that we can neglect at this point.
Nevertheless, these kinetic terms will become important in the presence of an external drive, which we discuss in Section \ref{sec:SOC}.
The validity of approximating the Hubbard model in Eqs.~\eqref{eq:H_kin_FH}--\eqref{eq:H_int_FH} by the spin model in Eqs.~\eqref{H_kin_start}--\eqref{eq:H_int_start} has been previously benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}, and we provide additional benchmarking for SU(4) and SU(6) in Appendix \ref{sec:benchmarking}.

To further simplify the interaction Hamiltonian $\H_{\t{int}}$ and write it in a form reminiscent of more familiar SU($2$) spin models, we now construct the operator-valued spin matrix
\begin{align}
  \v\s_q \equiv \sum_{\mu,\nu} \s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v{\hat{A}},\v{\hat{B}}$, we define the inner product
\begin{align}
  \v{\hat{A}} \cdot \v{\hat{B}}
  \equiv \sum_{\mu,\nu} \hat A_{\mu\nu}^\dag \hat B_{\mu\nu}.
  \label{eq:dot_product}
\end{align}
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  \H_{\t{int}} = -\f{u}{2N} \sum_{p,q} \v\s_p\cdot\v\s_q
  = -\f{u}{2N}\v\S\cdot\v\S,
  \label{eq:H_int}
\end{align}
where $\v\S\equiv\sum_q\v\s_q$ is a collective spin matrix, analogous to the collective spin vector $\vec S=(\S_\x,\S_\y,\S_\z)$ in the case of SU($2$) \cite{he2019engineering}, with $\frac12\v\S\cdot\v\S \simeq \vec S\cdot\vec S = \S_\x^2 + \S_\y^2 + \S_\z^2$ when $n=2$ (here $\simeq$ denotes equality up to identity terms).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{spin_model.pdf}
\caption{
(a) Ultracold atoms on a lattice of $L$ sites tunnel between neighboring lattice sites at a rate $J$, and locally repel each other with interaction energy $U$.
(b,c) When the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, the frozen-mode approximation enables the interaction Hamiltonian to be written as a spin model consisting of exchange terms $\v\s_p\cdot\v\s_q$, which swap the states two spins pinned to momentum modes $p,q$.
(d) Interactions open an energy gap $u=U\times N/L$ between the manifold of permutationally symmetric states of $N$ spins, and the orthogonal complement of states that break spin-permutation symmetry.
}
\label{fig:spin_model}
\end{figure}

We now discuss the spin Hamiltonian $\H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v\s_p\cdot\v\s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v\s_p\cdot\v\s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v\s_p\cdot\v\s_q$ is analogous to the enforcement of SU($2$) spin alignment by ferromagnetic interactions, which similarly assigns different energies to the anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $-\v\s_p\cdot\v\s_q$, the interaction Hamiltonian $\H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $u$ between the manifold of all permutationally symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

In the case of SU($2$), the PS manifold is precisely the Dicke manifold of collective states $\ket{m_\z}$ with total spin $S=\frac{N}{2}$ and definite spin projection $m_\z\in\set{S,S-1,\cdots,-S}$ onto a fixed quantization axis.
Equivalently, Dicke states $\ket{m_\z}=\ket{m_\up,m_\dn}$ can be labeled by a definite number of spins $m_\up=S+m_\z$ ($m_\dn=S-m_\z$) pointing up (down) along the spin quantization axis, with $m_\up+m_\dn=N$.
In the general case of SU($n$), the PS manifold is similarly spanned by states $\ket{m_s,m_{s-1},\cdots,m_{-s}}$ with a definite number $m_\mu$ of spins in state $\mu$, and $\sum_\mu m_\mu=N$.
The dimension of the PS manifold is equal to the number of ways of assigning $N$ identical spins to $n$ distinct internal states, or ${N+n-1 \choose n-1} \sim N^{n-1}$.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{qudit_rep.pdf}
\caption{
Whereas the state of a two-level spin (qubit) can be represented by a point on (or inside) the Bloch sphere, the state of a $n$-level spin is more generally represented by a probability distribution on the sphere.
The distribution shown for $n=10$ corresponds to a Haar-random pure state.
}
\label{fig:spin_dist}
\end{figure}

Finally, we take a moment to discuss individual $n$-level spins.
The state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of a $n$-level spin can be represented by a quasi-probability distribution $Q_\psi$ on the sphere (commonly known as the Husimi-$Q$ function, e.g.~in the spin-squeezing community \cite{ma2011quantum}).
The value $Q_\psi\p{\v v}$ at a point $\v v$ on the sphere is equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ maximally polarized in the direction of $\v v$: $Q_\psi\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\hat\rho$, this distribution is defined by $Q_{\hat\rho}\equiv\bk{\v v|\hat\rho|\v v}$.
Closely related spherical representations of multilevel spin states and operators are discussed in Refs.~\cite{dowling1994wigner, li2013weylwignermoyal}.
In practice, it is conceptually useful to identify the Hilbert space of a single $n$-level spin with the Dicke manifold of $n-1$ spin-$\frac12$ particles.

@@

We consider a specific three-laser driving scheme with a geometry sketched in Figure \ref{fig:3LD}.
Here the lattice lies in the $y$-$z$ plane at an angle $\theta$ to the $z$ axis, oriented along $\v\ell=(0,\sin\theta,\cos\theta)$.
We set the spin quantization axis along $z$.
The laser setup consists of
\begin{enumerate*}
\item two counter-propagating right-circularly polarized lasers with with drive amplitudes $\Omega_\pm$ and wavevectors $\kappa\v v_\pm$, propagating in opposite directions along the $z$ axis, $\v v_\pm=(0,0,\pm1)$, and
\item a third laser linearly polarized along $z$, with drive amplitude $\Omega_0$ and wavevector $\kappa \v v_0$, propagating along the $x$ axis $\v v_0=(1,0,0)$.
\end{enumerate*}
All driving lasers are detuned by $\Delta$ below an electronic transition.
The full Hamiltonian for this three-laser drive can be written as
\begin{align}
  \H_{\t{3LD}}^{\t{full}}
  = \sum_{j,m} \Omega_m
  \p{e^{-\i m\phi j} \s_{mj} \otimes\op{\e}{\g}_j + \t{h.c.}}
  + \Delta \hat N_\e,
\end{align}
where $m\in\set{+1,0,-1}$ indexes the laser pointing along $\v v_m$; the SOC angle $\phi\equiv\kappa\v v_+\cdot\v\ell=\kappa a \cos\theta$ (with $a$ the lattice spacing); $\s_{\z,j},\s_{+,j}\s_{-,j}$ are standard axial, spin-raising, and spin-lowering operators for spin $j$; $\s_{0,j}\equiv\s_{\z,j}$ for shorthand; $\ket\g_j$ and $\ket\e_j$ respectively denote the ground and excited electronic states of atom $j$; and $\hat N_\e=\1\otimes\sum_j\op\e_j$ counts the number of excited atoms (with $\1$ the identity operator on all spin degrees of freedom).

@@

We thank V.~Gurarie and E.~Yuzbashyan for useful discussions on the topic.
This work was supported by the AFOSR grants FA9550-18-1-0319, FA9550-19-1-027, by the DARPA and ARO grant W911NF-16-1-0576, ARO W911NF-19-1-0210, DOE (QSA), NSF PHY1820885, NSF JILA-PFC PHY-1734006 and NSF QLCI-2016244 grants, and by NIST.

>>>>

\date{22 September 2021}

\begin{abstract}
We study multilevel fermions in an optical lattice described by the Hubbard model with on site SU($n$)-symmetric interactions.
We show that in an appropriate parameter regime this system can be mapped onto a spin model with all-to-all SU($n$)-symmetric couplings.
Raman pulses that address internal spin states modify the atomic dispersion relation and induce spin-orbit coupling, which can act as a synthetic inhomogeneous magnetic field that competes with the SU($n$) exchange interactions.
We investigate the mean-field dynamical phase diagram of the resulting model as a function of $n$ and different initial configurations that are accessible with Raman pulses.
Consistent with previous studies for $n=2$, we find that for some initial states the spin model exhibits two distinct dynamical phases that obey simple scaling relations with $n$.
Moreover, for $n>2$ we find that dynamical behavior can be highly sensitive to initial intra-spin coherences.
Our predictions are readily testable in current experiments with ultracold alkaline-earth(-like) atoms.
\end{abstract}

@@

SU($n$) symmetries play an important role in physics.
Underpinning much of high energy physics, the SU($n$) gauge theory known as Yang-Mills theory is central to our understanding of the electroweak and strong forces.
Extensions of Yang-Mills and SU($n$) symmetry feature in the most well-studied examples of holographic duality \cite{maldacena1999largen} and the connection between entanglement and gravity \cite{ryu2006holographic} through the anti-de Sitter/conformal field theory (AdS/CFT) correspondence.
In a condensed matter setting, SU(2) appears ubiquitously as a symmetry of the Hubbard model, with important consequences for the study of quantum magnetism and high temperature superconductivity \cite{lee2006doping}.
The extension of SU(2) Hubbard and spin models to SU($n$) has led to predictions of exotic phases of matter such as valence bond solids \cite{read1989valencebond, rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and chiral spin liquids \cite{hermele2009mott, hermele2011topological, chen2016syntheticgaugefield, nataf2016chiral}, as well as the potential to perform universal topological quantum computation \cite{freedman2004class, nayak2008nonabelian} and other phenomena \cite{nataf2014exact, nataf2016exact}.
Furthermore, disordered SU($n$) spin models have opened analytically tractable avenues for studying quantum chaos and information scrambling \cite{sachdev1993gapless}.

@@

In the spirit of quantum simulation, further investigations in controlled settings will play an important role in understanding the consequence of SU($n$) symmetries for fundamental questions in physics, as well as their practical use in technological applications.
For example, SU(2)-symmetric spin interactions can be harnessed to develop quantum sensors that surpass classical limits on measurement precision \cite{he2019engineering, perlin2020spin}.
The prospect of similarly exploiting more general SU($n$) symmetries to achieve a technological advantage is still an unexplored avenue of research with untapped potential.

In this work, we consider an experimentally relevant and theoretically tractable regime of the SU($n$) Hubbard model, highlighting differences and similarities with the more familiar case of SU(2).
Working at ultracold temperatures and unit spatial filling (one atom per lattice site), we begin by mapping the SU($n$) Hubbard model onto a multilevel spin model with all-to-all SU($n$)-symmetric interactions in Section \ref{sec:spin_model}.
In Section \ref{sec:controls} we consider the use of control fields to address nuclear spins, finding a simple three-laser driving scheme that allows for the preparation of interesting states with nontrivial intra-spin correlations when $n>2$.
We consider the effect of spin-orbit coupling (SOC) induced by control fields in Section \ref{sec:SOC}, finding in particular that the weak-SOC limit generally gives rise to a (synthetic) inhomogeneous magnetic field, extending previously known results to $n>2$ \cite{mancini2015observation, wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
Finally, we combine these ingredients to examine mean-field dynamical behaviors of the SU($n$) spin model in Section \ref{sec:mean_field}, finding that:
\begin{enumerate*}
\item long-time-averaged observables obey simple scaling relations with the spin dimension $n$, exhibiting (for spin-polarized initial states) dynamical ferromagnetic and dynamical paramagnetic phases, as previously seen for the case of $n=2$ \cite{smale2019observation, lewis-swan2021cavityqed}, and
\item for $n>2$ the long-time dynamics can be highly sensitive to the intra-spin coherences of the initial state.
\end{enumerate*}
We conclude and discuss future directions in Section \ref{sec:conclusions}.

@@

Here we derive a collective SU($n$) spin model for a system of ultracold alkaline-earth(-like) atoms trapped in an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by the single-body kinetic and two-body interaction Hamiltonians
\begin{align}
  \H_{\t{kin}}
  &= -J \sum_{\bk{j,j'},\mu} \c_{j\mu}^\dag \c_{j'\mu} + \t{h.c.},
  \label{eq:H_kin_FH} \\
  \H_{\t{int}}
  &= \f{U}{2} \sum_{j,\mu,\nu}
  \c_{j\mu}^\dag \c_{j\mu} \c_{j\nu}^\dag \c_{j\nu},
  \label{eq:H_int_FH}
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $\c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude (for simplicity assumed to be the same in all directions); and $U$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which is a good approximation for a sufficiently deep lattice, namely when $J\lesssim E_{\t{R}}$, where $E_{\t{R}}$ is the atom recoil energy.
For simplicity, we now assume a one-dimensional periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$ (in units with lattice spacing $a=1$), $\c_{j\mu} = \frac1{\sqrt{L}} \sum_q e^{-\i q\cdot j} \c_{q\mu}$, finding that
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \c_{q\mu}^\dag \c_{q\mu}, \\
  \H_{\t{int}}
  &= \f{u}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  \c_{k\mu}^\dag \c_{\ell\mu} \c_{p\nu}^\dag \c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $u\equiv U\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum).

If the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, then the mode-changing collisions in $\H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (i.e.~either $k=\ell$ and $p=q$, or $k=q$ and $p=\ell$)\footnote{Note that the frozen-mode approximation neglects correlated momentum-hopping terms of the form $\c_{\pi-p,\mu}^\dag \c_{\pi-q,\mu} \c_{p\nu}^\dag \c_{q\nu}$, which conserve both momentum and energy.
We defer a careful treatment of these terms to future work, noting only that they vanish on the manifold of permutationally symmetric spin states with one atom per lattice site, and that the frozen-mode approximation is benchmarked in Refs.~\cite{he2019engineering, smale2019observation} and Appendix \ref{sec:benchmarking}.}.
The terms with $k=\ell$ and $p=q$ are $\frac{u}{2N} \sum \c_{k\mu}^\dag \c_{k\mu} \c_{p\nu}^\dag \c_{p\nu} = \frac12 N u$, which is a constant  energy shift that we can freely neglect.
Defining the spin operators $\s_{\mu\nu q}\equiv \c_{q\mu}^\dag \c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians are
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \s_{\mu\mu q},
  \label{H_kin_start} \\
  \H_{\t{int}}
  &= -\f{u}{2N} \sum_{p,q,\mu,\nu} \s_{\mu\nu p} \s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case we can simply treat our system as $N$ distinguishable $n$-level quantum spins at ``lattice sites'' $p,q$.
Note that the ``kinetic'' terms of this spin model ($\H_{\t{kin}}$) are proportional to the identity operator, contributing an overall shift in energy that we can neglect at this point.
Nevertheless, these kinetic terms will become important in the presence of an external drive, which we discuss in Section \ref{sec:SOC}.
The validity of approximating the Hubbard model in Eqs.~\eqref{eq:H_kin_FH}--\eqref{eq:H_int_FH} by the spin model in Eqs.~\eqref{H_kin_start}--\eqref{eq:H_int_start} has been previously benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}, and we provide additional benchmarking for SU(4) and SU(6) in Appendix \ref{sec:benchmarking}.

To further simplify the interaction Hamiltonian $\H_{\t{int}}$ and write it in a form reminiscent of more familiar SU(2) spin models, we now construct the operator-valued spin matrix
\begin{align}
  \v\s_q \equiv \sum_{\mu,\nu} \s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v{\hat{A}},\v{\hat{B}}$, we define the inner product
\begin{align}
  \v{\hat{A}} \cdot \v{\hat{B}}
  \equiv \sum_{\mu,\nu} \hat A_{\mu\nu}^\dag \hat B_{\mu\nu}.
  \label{eq:dot_product}
\end{align}
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  \H_{\t{int}} = -\f{u}{2N} \sum_{p,q} \v\s_p\cdot\v\s_q
  = -\f{u}{2N}\v\S\cdot\v\S,
  \label{eq:H_int}
\end{align}
where $\v\S\equiv\sum_q\v\s_q$ is a collective spin matrix, analogous to the collective spin vector $\vec S=(\S_\x,\S_\y,\S_\z)$ in the case of SU(2) \cite{he2019engineering}, with $\frac12\v\S\cdot\v\S \simeq \vec S\cdot\vec S = \S_\x^2 + \S_\y^2 + \S_\z^2$ when $n=2$ (here $\simeq$ denotes equality up to identity terms).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{spin_model.pdf}
\caption{
(a) Ultracold atoms on a lattice of $L$ sites tunnel between neighboring lattice sites at a rate $J$, and locally repel each other with interaction energy $U$.
(b,c) When the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, the frozen-mode approximation enables the interaction Hamiltonian to be written as a spin model consisting of exchange terms $\v\s_p\cdot\v\s_q$, which swap the states of two spins pinned to modes $p,q$.
(d) Interactions open an energy gap $u=U\times N/L$ between the manifold of permutationally symmetric states of $N$ spins, and the orthogonal complement of states that break spin-permutation symmetry.
}
\label{fig:spin_model}
\end{figure}

We now discuss the spin Hamiltonian $\H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v\s_p\cdot\v\s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v\s_p\cdot\v\s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v\s_p\cdot\v\s_q$ is analogous to the enforcement of SU(2) spin alignment by ferromagnetic interactions, which similarly assigns different energies to the anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $\v\s_p\cdot\v\s_q$, the interaction Hamiltonian $\H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $u$ between the manifold of all permutationally symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

In the case of SU(2), the PS manifold is precisely the Dicke manifold of collective states $\ket{m_\z}$ with total spin $S=\frac{N}{2}$ and definite spin projection $m_\z\in\set{S,S-1,\cdots,-S}$ onto a fixed quantization axis.
Equivalently, Dicke states $\ket{m_\z}=\ket{m_\up,m_\dn}$ can be labeled by a definite number of spins $m_\up=S+m_\z$ ($m_\dn=S-m_\z$) pointing up (down) along the spin quantization axis, with $m_\up+m_\dn=N$.
In the general case of SU($n$), the PS manifold is similarly spanned by states $\ket{m_s,m_{s-1},\cdots,m_{-s}}$ with a definite number $m_\mu$ of spins in state $\mu$, and $\sum_\mu m_\mu=N$.
The dimension of the PS manifold is equal to the number of ways of assigning $N$ identical spins to $n$ distinct internal states, or ${N+n-1 \choose n-1} \sim N^{n-1}$.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{qudit_rep.pdf}
\caption{
Whereas the state of a two-level spin (qubit) can be represented by a point on (or inside) the Bloch sphere, the state of an $n$-level spin is more generally represented by a probability distribution on the Bloch sphere.
The distribution shown for $n=10$ corresponds to a Haar-random pure state.
}
\label{fig:spin_dist}
\end{figure}

Finally, we take a moment to discuss individual $n$-level spins.
The state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of an $n$-level spin can be represented by a quasi-probability distribution $Q_\psi$ on the Bloch sphere (commonly known as the Husimi-$Q$ function, e.g.~in the spin-squeezing community \cite{ma2011quantum}).
The value $Q_\psi\p{\v v}$ at a point $\v v$ on the sphere is equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ that is maximally polarized in the direction of $\v v$: $Q_\psi\p{\v v}\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\hat\rho$, this distribution is defined by $Q_{\hat\rho}\p{\v v}\equiv\bk{\v v|\hat\rho|\v v}$.
Closely related spherical representations of multilevel spin states and operators are discussed in Refs.~\cite{dowling1994wigner, li2013weylwignermoyal}.
In practice, it is conceptually useful to identify the Hilbert space of a single $n$-level spin with the Dicke manifold of $n-1$ spin-$\frac12$ particles.

@@

We consider a specific three-laser driving scheme with a geometry sketched in Figure \ref{fig:3LD}.
Here the lattice lies in the $y$-$z$ plane at an angle $\theta$ to the $z$ axis, oriented along $\v\ell=(0,\sin\theta,\cos\theta)$.
We set the spin quantization axis along $z$.
The laser setup consists of
\begin{enumerate*}
\item two counter-propagating right-circularly polarized lasers with drive amplitudes $\Omega_\pm$ and wavevectors $\kappa\v v_\pm$, propagating in opposite directions along the $z$ axis, $\v v_\pm=(0,0,\pm1)$, and
\item a third laser linearly polarized along $z$, with drive amplitude $\Omega_0$ and wavevector $\kappa \v v_0$, propagating along the $x$ axis $\v v_0=(1,0,0)$.
\end{enumerate*}
All driving lasers are detuned by $\Delta$ below an electronic transition.
The full Hamiltonian for this three-laser drive can be written as
\begin{align}
  \H_{\t{3LD}}^{\t{full}}
  = \sum_{j,m} \Omega_m
  \p{e^{-\i m\phi j} \s_{mj} \otimes\op{\e}{\g}_j + \t{h.c.}}
  + \Delta \hat N_\e,
\end{align}
where $m\in\set{+1,0,-1}$ indexes the laser pointing along $\v v_m$; the SOC angle $\phi\equiv\kappa\v v_+\cdot\v\ell=\kappa \cos\theta$ (in units with lattice spacing $a=1$); $\s_{\z,j},\s_{+,j}\s_{-,j}$ are standard axial, spin-raising, and spin-lowering operators for the spin at lattice site $j$; $\s_{0,j}\equiv\s_{\z,j}$ for shorthand; $\ket\g_j$ and $\ket\e_j$ respectively denote the ground and excited electronic states of atom $j$; and $\hat N_\e=\1\otimes\sum_j\op\e_j$ counts the number of excited atoms (with $\1$ the identity operator on all spin degrees of freedom).

@@

We thank Victor Gurarie, Emil Yuzbashyan, Asier P.~Orioli, and Jeremy T.~Young for helpful discussions on this work.
This work was supported by the AFOSR grants FA9550-18-1-0319, FA9550-19-1-027, by the DARPA and ARO grant W911NF-16-1-0576, ARO W911NF-19-1-0210, DOE (QSA), NSF PHY1820885, NSF JILA-PFC PHY-1734006, NSF QLCI-2016244, and by NIST.
